---
title: "Measuring Value: A Quantitative Study of NBA Performance Trends"
author: "Sejal Dhanta"
date: "2026-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This assignment explores a decade of professional basketball performance data (2005–2014) for ten elite NBA players. The objective of this analysis is to understand the relationship between player performance, salary growth, scoring efficiency, and consistency over time using R programming.

In professional sports, performance drives market value. However, financial rewards are influenced not only by current performance but also by reputation, consistency, contract timing, and strategic positioning. Through statistical modeling and visualization, this assignment attempts to answer key business questions:

Does higher performance guarantee higher salary?

Are salaries aligned with scoring output?

Which players show long-term consistency?

How does market value evolve over time?

This project combines data structuring, matrix manipulation, statistical computation, and visualization techniques to derive meaningful managerial insights.

## Data loading and matrix creation

The dataset includes:

Seasons (2005–2014)

Players (10 elite athletes)

Salary

Games Played

Minutes Played

Field Goals

Field Goal Attempts

Total Points

##### Why Matrix Structure?

Using matrices allows:

Structured row-wise player comparison

Column-wise season comparison

Efficient mathematical operations

Clean labeling using rownames and colnames

Rows represent players
Columns represent seasons

This structure enables multi-dimensional performance tracking across time.

#### Seasons
```{r}
Seasons <- c("2005","2006","2007","2008","2009","2010","2011","2012","2013","2014")
```
#### Players
```{r}
Players <- c("KobeBryant","JoeJohnson","LeBronJames","CarmeloAnthony","DwightHoward","ChrisBosh","ChrisPaul","KevinDurant","DerrickRose","DwayneWade")
```
#### Salaries
```{r}
KobeBryant_Salary <- c(15946875,17718750,19490625,21262500,23034375,24806250,25244493,27849149,30453805,23500000)
JoeJohnson_Salary <- c(12000000,12744189,13488377,14232567,14976754,16324500,18038573,19752645,21466718,23180790)
LeBronJames_Salary <- c(4621800,5828090,13041250,14410581,15779912,14500000,16022500,17545000,19067500,20644400)
CarmeloAnthony_Salary <- c(3713640,4694041,13041250,14410581,15779912,17149243,18518574,19450000,22407474,22458000)
DwightHoward_Salary <- c(4493160,4806720,6061274,13758000,15202590,16647180,18091770,19536360,20513178,21436271)
ChrisBosh_Salary <- c(3348000,4235220,12455000,14410581,15779912,14500000,16022500,17545000,19067500,20644400)
ChrisPaul_Salary <- c(3144240,3380160,3615960,4574189,13520500,14940153,16359805,17779458,18668431,20068563)
KevinDurant_Salary <- c(0,0,4171200,4484040,4796880,6053663,15506632,16669630,17832627,18995624)
DerrickRose_Salary <- c(0,0,0,4822800,5184480,5546160,6993708,16402500,17632688,18862875)
DwayneWade_Salary <- c(3031920,3841443,13041250,14410581,15779912,14200000,15691000,17182000,18673000,15000000)
```
#### Matrix
```{r}
Salary <- rbind(KobeBryant_Salary, JoeJohnson_Salary, LeBronJames_Salary, CarmeloAnthony_Salary, DwightHoward_Salary, ChrisBosh_Salary, ChrisPaul_Salary, KevinDurant_Salary, DerrickRose_Salary, DwayneWade_Salary)
Salary
rm(KobeBryant_Salary, JoeJohnson_Salary, CarmeloAnthony_Salary, DwightHoward_Salary, ChrisBosh_Salary, LeBronJames_Salary, ChrisPaul_Salary, DerrickRose_Salary, DwayneWade_Salary, KevinDurant_Salary)
colnames(Salary) <- Seasons
rownames(Salary) <- Players
```
#### Games 
```{r}
KobeBryant_G <- c(80,77,82,82,73,82,58,78,6,35)
JoeJohnson_G <- c(82,57,82,79,76,72,60,72,79,80)
LeBronJames_G <- c(79,78,75,81,76,79,62,76,77,69)
CarmeloAnthony_G <- c(80,65,77,66,69,77,55,67,77,40)
DwightHoward_G <- c(82,82,82,79,82,78,54,76,71,41)
ChrisBosh_G <- c(70,69,67,77,70,77,57,74,79,44)
ChrisPaul_G <- c(78,64,80,78,45,80,60,70,62,82)
KevinDurant_G <- c(35,35,80,74,82,78,66,81,81,27)
DerrickRose_G <- c(40,40,40,81,78,81,39,0,10,51)
DwayneWade_G <- c(75,51,51,79,77,76,49,69,54,62)
```
#### Matrix
```{r}
Games <- rbind(KobeBryant_G, JoeJohnson_G, LeBronJames_G, CarmeloAnthony_G, DwightHoward_G, ChrisBosh_G, ChrisPaul_G, KevinDurant_G, DerrickRose_G, DwayneWade_G)
rm(KobeBryant_G, JoeJohnson_G, CarmeloAnthony_G, DwightHoward_G, ChrisBosh_G, LeBronJames_G, ChrisPaul_G, DerrickRose_G, DwayneWade_G, KevinDurant_G)
colnames(Games) <- Seasons
rownames(Games) <- Players
```
#### Minutes Played
```{r}
KobeBryant_MP <- c(3277,3140,3192,2960,2835,2779,2232,3013,177,1207)
JoeJohnson_MP <- c(3340,2359,3343,3124,2886,2554,2127,2642,2575,2791)
LeBronJames_MP <- c(3361,3190,3027,3054,2966,3063,2326,2877,2902,2493)
CarmeloAnthony_MP <- c(2941,2486,2806,2277,2634,2751,1876,2482,2982,1428)
DwightHoward_MP <- c(3021,3023,3088,2821,2843,2935,2070,2722,2396,1223)
ChrisBosh_MP <- c(2751,2658,2425,2928,2526,2795,2007,2454,2531,1556)
ChrisPaul_MP <- c(2808,2353,3006,3002,1712,2880,2181,2335,2171,2857)
KevinDurant_MP <- c(1255,1255,2768,2885,3239,3038,2546,3119,3122,913)
DerrickRose_MP <- c(1168,1168,1168,3000,2871,3026,1375,0,311,1530)
DwayneWade_MP <- c(2892,1931,1954,3048,2792,2823,1625,2391,1775,1971)
```
#### Matrix
```{r}
MinutesPlayed <- rbind(KobeBryant_MP, JoeJohnson_MP, LeBronJames_MP, CarmeloAnthony_MP, DwightHoward_MP, ChrisBosh_MP, ChrisPaul_MP, KevinDurant_MP, DerrickRose_MP, DwayneWade_MP)
rm(KobeBryant_MP, JoeJohnson_MP, CarmeloAnthony_MP, DwightHoward_MP, ChrisBosh_MP, LeBronJames_MP, ChrisPaul_MP, DerrickRose_MP, DwayneWade_MP, KevinDurant_MP)
colnames(MinutesPlayed) <- Seasons
rownames(MinutesPlayed) <- Players
```
#### Field Goals
```{r}
KobeBryant_FG <- c(978,813,775,800,716,740,574,738,31,266)
JoeJohnson_FG <- c(632,536,647,620,635,514,423,445,462,446)
LeBronJames_FG <- c(875,772,794,789,768,758,621,765,767,624)
CarmeloAnthony_FG <- c(756,691,728,535,688,684,441,669,743,358)
DwightHoward_FG <- c(468,526,583,560,510,619,416,470,473,251)
ChrisBosh_FG <- c(549,543,507,615,600,524,393,485,492,343)
ChrisPaul_FG <- c(407,381,630,631,314,430,425,412,406,568)
KevinDurant_FG <- c(306,306,587,661,794,711,643,731,849,238)
DerrickRose_FG <- c(208,208,208,574,672,711,302,0,58,338)
DwayneWade_FG <- c(699,472,439,854,719,692,416,569,415,509)
```
#### Matrix

```{r}
FieldGoals <- rbind(KobeBryant_FG, JoeJohnson_FG, LeBronJames_FG, CarmeloAnthony_FG, DwightHoward_FG, ChrisBosh_FG, ChrisPaul_FG, KevinDurant_FG, DerrickRose_FG, DwayneWade_FG)
rm(KobeBryant_FG, JoeJohnson_FG, LeBronJames_FG, CarmeloAnthony_FG, DwightHoward_FG, ChrisBosh_FG, ChrisPaul_FG, KevinDurant_FG, DerrickRose_FG, DwayneWade_FG)
colnames(FieldGoals) <- Seasons
rownames(FieldGoals) <- Players
```
#### Field Goal Attempts
```{r}
KobeBryant_FGA <- c(2173,1757,1690,1712,1569,1639,1336,1595,73,713)
JoeJohnson_FGA <- c(1395,1139,1497,1420,1386,1161,931,1052,1018,1025)
LeBronJames_FGA <- c(1823,1621,1642,1613,1528,1485,1169,1354,1353,1279)
CarmeloAnthony_FGA <- c(1572,1453,1481,1207,1502,1503,1025,1489,1643,806)
DwightHoward_FGA <- c(881,873,974,979,834,1044,726,813,800,423)
ChrisBosh_FGA <- c(1087,1094,1027,1263,1158,1056,807,907,953,745)
ChrisPaul_FGA <- c(947,871,1291,1255,637,928,890,856,870,1170)
KevinDurant_FGA <- c(647,647,1366,1390,1668,1538,1297,1433,1688,467)
DerrickRose_FGA <- c(436,436,436,1208,1373,1597,695,0,164,835)
DwayneWade_FGA <- c(1413,962,937,1739,1511,1384,837,1093,761,1084)
```
#### Matrix

```{r}
FieldGoalAttempts <- rbind(KobeBryant_FGA, JoeJohnson_FGA, LeBronJames_FGA, CarmeloAnthony_FGA, DwightHoward_FGA, ChrisBosh_FGA, ChrisPaul_FGA, KevinDurant_FGA, DerrickRose_FGA, DwayneWade_FGA)
rm(KobeBryant_FGA, JoeJohnson_FGA, LeBronJames_FGA, CarmeloAnthony_FGA, DwightHoward_FGA, ChrisBosh_FGA, ChrisPaul_FGA, KevinDurant_FGA, DerrickRose_FGA, DwayneWade_FGA)
colnames(FieldGoalAttempts) <- Seasons
rownames(FieldGoalAttempts) <- Players
```
#### Points
```{r}
KobeBryant_PTS <- c(2832,2430,2323,2201,1970,2078,1616,2133,83,782)
JoeJohnson_PTS <- c(1653,1426,1779,1688,1619,1312,1129,1170,1245,1154)
LeBronJames_PTS <- c(2478,2132,2250,2304,2258,2111,1683,2036,2089,1743)
CarmeloAnthony_PTS <- c(2122,1881,1978,1504,1943,1970,1245,1920,2112,966)
DwightHoward_PTS <- c(1292,1443,1695,1624,1503,1784,1113,1296,1297,646)
ChrisBosh_PTS <- c(1572,1561,1496,1746,1678,1438,1025,1232,1281,928)
ChrisPaul_PTS <- c(1258,1104,1684,1781,841,1268,1189,1186,1185,1564)
KevinDurant_PTS <- c(903,903,1624,1871,2472,2161,1850,2280,2593,686)
DerrickRose_PTS <- c(597,597,597,1361,1619,2026,852,0,159,904)
DwayneWade_PTS <- c(2040,1397,1254,2386,2045,1941,1082,1463,1028,1331)
```
#### Matrix
```{r}
Points <- rbind(KobeBryant_PTS, JoeJohnson_PTS, LeBronJames_PTS, CarmeloAnthony_PTS, DwightHoward_PTS, ChrisBosh_PTS, ChrisPaul_PTS, KevinDurant_PTS, DerrickRose_PTS, DwayneWade_PTS)
rm(KobeBryant_PTS, JoeJohnson_PTS, LeBronJames_PTS, CarmeloAnthony_PTS, DwightHoward_PTS, ChrisBosh_PTS, ChrisPaul_PTS, KevinDurant_PTS, DerrickRose_PTS, DwayneWade_PTS)
colnames(Points) <- Seasons
rownames(Points) <- Players
```
#### To visualize the FieldGoals scored by each player  
```{r}
matplot(t(FieldGoals), type = "b", pch = 15:18, col = c(1:4, 6))
legend("bottomleft", inset = 0.01, legend = Players, c(1:4, 6), pch = 15:18, horiz = F)
```
```{r}
matplot(t(FieldGoals/Games), type = "b", pch = 15:18, col = c(1:4, 6))
legend("bottomleft", inset = 0.01, legend = Players, c(1:4, 6), pch = 15:18, horiz = F)
```

#### To check the accuracy of the shots for each player visualization 
```{r}
matplot(t(FieldGoals/FieldGoalAttempts), type = "b", pch = 15:18, col = c(1:4, 6))
legend("bottomleft", inset = 0.01, legend = Players, c(1:4, 6), pch = 15:18, horiz = F)
```


# Convert to dataframe
```{r}
data <- data.frame(
  Player = rep(Players, each=10),
  Season = rep(Seasons, times=10),
  Salary = as.vector(t(Salary)),
  Points = as.vector(t(Points)),
  Games = as.vector(t(Games))
)
```
```{r}
data$PPG <- data$Points / data$Games
```

# 📈 Visualisations and Interpretations

---

## 1. Player Performance Trend
```{r}
library(ggplot2)
```

```{r}
ggplot(data, aes(Season, PPG, color=Player, group=Player)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(title="PPG Trend Over Seasons")
```

#### Insights

1️⃣ Kobe peaked highest in early seasons

2️⃣ Durant shows strong upward trajectory

3️⃣ LeBron maintains consistent elite performance

4️⃣ Derrick Rose displays performance volatility

5️⃣ Post-2012 overall scoring decline observed

6️⃣ 2008–2012 highly competitive scoring era

7️⃣ Few players show long-term stability

---

## 2. Salary Trend

```{r}
ggplot(data, aes(Season, Salary/1000000, color=Player, group=Player)) +
  geom_line(linewidth=1) +
  labs(title="Salary Trend (Millions)")
```

#### Insights

1️⃣ Overall player salaries show consistent upward growth trend.

2️⃣ Kobe Bryant remained highest paid across most seasons.

3️⃣ Significant salary peaks observed around 2012–2013 period.

4️⃣ Kevin Durant experienced sharp contract-based salary escalation.

5️⃣ Derrick Rose saw major increase after performance peak.

6️⃣ League-wide salary inflation visible over entire timeline.

7️⃣ Few players experienced minor salary correction in 2014.

8️⃣ Salary growth generally aligns with star player status.

---

## 3. Salary vs Performance

```{r}
ggplot(subset(data, is.finite(PPG)), 
       aes(Salary/1000000, PPG, color = Player)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  labs(title = "Salary vs Performance Relationship",
       x = "Salary (Million $)",
       y = "Points Per Game") +
  theme_minimal()
```

#### Insights


1️⃣ Weak overall correlation between salary and performance.

2️⃣ Higher salaries do not guarantee higher scoring.

3️⃣ Some players show declining performance with salary growth.

4️⃣ Kobe shows negative salary–performance relationship trend.

5️⃣ Durant reflects moderate positive salary–performance association.

6️⃣ Joe Johnson shows performance decline at higher salaries.

7️⃣ Mid-salary range displays clustered performance consistency.

8️⃣ Performance variability higher among top-paid players.

9️⃣ Market value not perfectly aligned with scoring output.

🔟 Salary increases sometimes follow past peak performance.

---

## 4. Salary Distribution

```{r}
ggplot(data, aes(Salary/1000000)) +
  geom_histogram(fill="steelblue", bins=15)
```

#### Insights

1️⃣ Salary distribution appears moderately right-skewed overall.

2️⃣ Majority salaries cluster between 12–20 million dollars.

3️⃣ Few extreme high-value salary outliers observed.

4️⃣ Lower salary range shows limited player concentration.

5️⃣ Distribution reflects widening compensation gap among players.

6️⃣ Mid-tier salary bracket dominates frequency counts.

7️⃣ High salaries represent elite market value players.

8️⃣ Overall spread indicates significant salary variability.

---

## 5. Scoring Distribution

```{r}
ggplot(subset(data, is.finite(PPG)), aes(PPG)) +
  geom_density(fill = "orange", alpha = 0.5) +
  labs(title = "Distribution of Points Per Game",
       x = "Points Per Game",
       y = "Density") +
  theme_minimal()

```

#### Insights


1️⃣ PPG distribution shows slight right-skewed tendency.

2️⃣ Majority performances cluster between 18–28 PPG.

3️⃣ Few extreme high-scoring outliers above 30 PPG.

4️⃣ Mid-range scoring dominates overall player performance.

5️⃣ Distribution indicates competitive scoring consistency across players.

6️⃣ Very low scoring seasons appear relatively rare.

7️⃣ Performance spread reflects moderate variability in scoring.

8️⃣ Peak density observed around 24–27 PPG range.

---

## 6. Top Performer Each Season

```{r}
library(dplyr)

top <- data %>%
  filter(is.finite(PPG)) %>%
  group_by(Season) %>%
  filter(PPG == max(PPG))

ggplot(top, aes(Season, PPG, fill = Player)) +
  geom_col() +
  theme_minimal()

```

#### Insights


1️⃣ Kobe Bryant records highest peak around 35 PPG.

2️⃣ Scoring leaders consistently exceed 27 PPG.

3️⃣ Kevin Durant leads multiple seasons consecutively.

4️⃣ Peak scoring fluctuates slightly across seasons.

5️⃣ 2005 shows highest single-season scoring performance.

6️⃣ Most seasons’ leaders score near 30 PPG.

7️⃣ Competitive dominance shifts among elite players.

8️⃣ No drastic scoring drop among seasonal leaders.

---

## 7. Average Salary Comparison

```{r}
avg_salary <- aggregate(Salary ~ Player, data, mean)

ggplot(avg_salary, aes(reorder(Player, Salary), Salary/1000000)) +
  geom_bar(stat="identity", fill="darkgreen") +
  coord_flip()
```

#### Insights

1️⃣ Kobe Bryant earns highest average salary overall.

2️⃣ Joe Johnson ranks second in salary earnings.

3️⃣ Carmelo Anthony and LeBron earn similarly.

4️⃣ Derrick Rose shows lowest average salary.

5️⃣ Significant salary gap between top and bottom.

6️⃣ Veteran stars command consistently higher salaries.

7️⃣ Salary ranking does not perfectly match scoring.

8️⃣ Financial rewards vary despite similar performance levels.

---

## 8. Boxplot of Salary by Player

```{r}
ggplot(data, aes(Player, Salary/1000000)) +
  geom_boxplot(fill="purple") +
  coord_flip()
```

#### Insights

1️⃣ Kobe Bryant shows highest median salary overall.

2️⃣ Derrick Rose displays wide salary variability.

3️⃣ Kevin Durant salary spread increases significantly.

4️⃣ LeBron James maintains consistently high earnings.

5️⃣ Several players exhibit lower-end salary outliers.

6️⃣ Joe Johnson salaries remain relatively stable.

7️⃣ Salary distributions vary notably among star players.

8️⃣ Top earners demonstrate higher upper salary ranges.

---

## 9. Boxplot of PPG by Player

```{r}
ggplot(subset(data, is.finite(PPG)), aes(Player, PPG)) +
  geom_boxplot(fill="red") +
  coord_flip() +
  labs(title="PPG Distribution by Player")
```


#### Insights


1️⃣ LeBron James shows consistently high scoring median.

2️⃣ Kobe Bryant exhibits high peak scoring outliers.

3️⃣ Kevin Durant maintains strong and stable scoring.

4️⃣ Carmelo Anthony records consistently high PPG range.

5️⃣ Derrick Rose shows wider performance variability.

6️⃣ Joe Johnson demonstrates moderate but stable scoring.

7️⃣ Dwyane Wade maintains solid mid-to-high scoring.

8️⃣ Elite scorers cluster above 24 PPG median.

---

## 10. Games vs Points

```{r}
ggplot(data, aes(Games, Points, color=Player)) +
  geom_point(size=3)
```

#### Insights

1️⃣ Positive relationship between salary and scoring output.

2️⃣ Higher scorers generally earn higher salaries.

3️⃣ Some players overperform relative to salary.

4️⃣ Regression line shows moderate upward trend.

5️⃣ Performance partly explains salary differences.

6️⃣ Few deviations from overall salary-performance trend.

7️⃣ Elite scorers cluster in higher salary range.

8️⃣ Salary not perfectly proportional to scoring.

---

## 11. Season-wise Average PPG

```{r}
season_avg <- aggregate(PPG ~ Season, data, mean)

ggplot(season_avg, aes(Season, PPG, group=1)) +
  geom_line(linewidth = 1, color="blue") +
  geom_point(size = 3)
```

#### Insights

1️⃣ Scoring distribution slightly right-skewed overall

2️⃣ Most players score between 18–28 PPG
.
3️⃣ High-scoring seasons remain relatively rare.

4️⃣ Density peak occurs around mid-20s PPG.

5️⃣ Extreme low performances are uncommon.

6️⃣ Scoring consistency visible across seasons.

7️⃣ Distribution shows moderate scoring variability.

8️⃣ Few standout elite scoring performances.

---

## 12. Performance Consistency (SD)

```{r}
sd_ppg <- aggregate(PPG ~ Player, data, sd)

ggplot(sd_ppg, aes(reorder(Player, PPG), PPG)) +
  geom_bar(stat="identity", fill="brown") +
  coord_flip()
```

#### Insights

1️⃣ Seasonal dominance shifts among elite players.

2️⃣ Kevin Durant leads multiple consecutive seasons.

3️⃣ Kobe Bryant records highest single-season peak.

4️⃣ Leaders consistently exceed 27 PPG.

5️⃣ Competitive scoring maintained across seasons.

6️⃣ No drastic decline among seasonal leaders.

7️⃣ Elite players rotate scoring leadership.

8️⃣ Performance levels remain consistently high.

## Strategic Insights

1️⃣ Salary growth reflects brand value and past reputation more than current season performance alone.

2️⃣ Consistent high performers create long-term franchise stability and reduce strategic investment risk.

3️⃣ Performance volatility significantly increases financial uncertainty in long-term player contracts.

4️⃣ Injury history must be incorporated into predictive salary and risk valuation models.

5️⃣ Market-wide salary inflation requires relative benchmarking rather than absolute compensation comparison.

6️⃣ Elite players command premium salaries due to revenue impact beyond statistical contribution.

7️⃣ Regression results indicate performance explains salary partially but not exclusively.

8️⃣ Efficient scorers generate higher return on investment with fewer resource inputs.

9️⃣ Contract timing often determines earnings more strongly than peak scoring output.

🔟 Data-driven evaluation reduces emotional bias in high-value contract negotiations.

1️⃣1️⃣ Top seasonal scorers influence competitive positioning but not always salary leadership.

1️⃣2️⃣ Balanced investment across consistent performers may outperform single-star dependency strategy.

## Conclusion

This assignment successfully integrates statistical analysis, visualization techniques, and managerial thinking to evaluate player performance and salary dynamics over a ten-year period.

The findings indicate that while performance influences compensation, salary structures are also shaped by reputation, contract timing, market inflation, and risk factors such as injuries. Elite players demonstrate both scoring dominance and financial advantage, yet consistency proves more valuable than short-term peaks.

The key takeaway is clear:

Sustainable competitive advantage comes from investing in consistent, efficient, and strategically valued talent not merely high-scoring stars.

This project reinforces the importance of combining data analytics with strategic business reasoning in decision-making environments.