---
title: "1st Assignment"
author: "Sejal Dhanta"
date: "2026-02-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction - R

R is based on an original set of notes describing the S and S-Plus
environments, written in 1990–1992 by Bill Venables and David M. Smith
while at the University of Adelaide.

R is an integrated suite of software facilities for data manipulation,
calculation and graphical display. Among other things it has

• an effective data handling and storage facility,

• a suite of operators for calculations on arrays, in particular
matrices,

• a large, coherent, integrated collection of intermediate tools for
data analysis,

• graphical facilities for data analysis, and

• a well-developed, simple, and effective programming language (called
‘S’) which includes conditionals, loops, user defined recursive
functions and input and output facilities.

### Vectors and assignment 

R operates on named data structures. The simplest such structure is the numeric vector, which 
is a single entity consisting of an ordered collection of numbers. To set up a vector named x, 
say, consisting of five numbers, namely 10.4, 5.6, 3.1, 6.4 and 21.7, use the R command 

```{r}
x  <- c(10.4, 5.6, 3.1, 6.4, 21.7)
```
This is an assignment statement using the function c() which in this context can take an 
arbitrary number of vector arguments and whose value is a vector got by concatenating its 
arguments end to end. 
Notice that the assignment operator (‘<-’), which consists of the two characters ‘<’ (“less 
than”) and ‘-’ (“minus”) occurring strictly side-by-side and it ‘points’ to the object receiving 
the value of the expression. 
The further assignment 
y <- c(x, 0, x) 
would create a vector y with 11 entries consisting of two copies of x with a zero in the middle 
place.

### Data permanency and removing objects
The entities that R creates and manipulates are known as objects. These may be variables, 
arrays of numbers, character strings, functions, or more general structures built from such 
components. 
During an R session, objects are created and stored by name. The collection of objects 
currently stored is called the workspace. 
To remove objects the function rm is available 

```{r,eval=FALSE}
rm(x, y, z, ink, junk, temp, foo) 
rm(x)
```

### Generating regular sequences 

R has a number of facilities for generating commonly used sequences of numbers. The 
function seq() is a more general facility for generating sequences. 
For example

```{r, eval=FALSE}
1:30 is the vector c(1, 2, ..., 29, 30). 
```

### Project 1: Mispriced Diamonds 

In the given data set, all rows represent Diamond with three columns namely carat (weight), 

clarity and price (price at which diamond is sold). 

More than 50,000 transactions are recorded in the file. High clarity Diamond is priced high!? 

Does the relationship between price and clarity always hold true? 
Let’s go and investigate! 

Step 1: Run the below line in Script

mydata <- read.csv(file.choose())

```{r}
mydata <- read.csv(file.choose()) 
```

It allows you to select your csv file and load data into RStudio 

Step 2: Install Package “ggplot2” by running below line in script. It helps to visualize the data 

```{r,eval=FALSE}
install.packages("ggplot2")
```
Once you run the program, your RStudio will connect to the internet and start downloading 
the said package for you. 
After installation of package, make sure you ‘activate’ ggplot2 under packages 

Step 3: Scatter Plot - Run the below line

```{r}
library(ggplot2)
ggplot(data=mydata, aes(x=carat, y=price)) +   geom_point()
```

This gives the scatter plot for Diamonds. All the records are illustrated in the plot. But there 
is no ‘clarity’ variable. 
This gives the scatter plot for Diamonds. All the records are illustrated in the plot. But there 
is no ‘clarity’ variable. 

Step 4: Now, you can add ‘clarity’ variable as color. Run below line

```{r}
ggplot(data=mydata, aes(x=carat, y=price,  color=clarity)) +  geom_point()
```

Step 5: The scatter plot has got many clumsy points and not so clear. To make it more clear 
run below line

```{r}
ggplot(data=mydata, aes(x=carat, y=price,  color=clarity)) +   geom_point(alpha=0.1) 
```

Step 6: It seems to be clear but scatter plot is having all the records irrespective of significant 
and not significant. Therefore, let’s filter out the data now with the condition on clarity. Run 
below line

```{r}
ggplot(data=mydata[mydata$carat < 2.5,], aes(x=carat, y=price, color=clarity)) + geom_point(alpha=0.1) 
```

Variable ‘carat’ records with only less than 2.5 values are considered in the above 
visualization. This is much clear! 
Step 7: To see the averages for a variable ‘carat’ run below line

```{r}
ggplot(data=mydata[mydata$carat < 2.5,],   aes(x=carat, y=price, color=clarity)) +  
geom_point(alpha=0.1) + geom_smooth() 
```

### Key Interpretation

#### 1. Strong positive relationship between carat and price

As carat size increases, the price of diamonds increases sharply.

The relationship is non-linear: price rises slowly at lower carats and accelerates after ~1 carat.

Interpretation: Larger diamonds command disproportionately higher prices.

#### 2. Clarity significantly affects price at the same carat

For the same carat value, diamonds with higher clarity grades (VVS1, VVS2, IF) are priced higher.

Lower clarity diamonds (especially I1) consistently lie at the bottom price range.

Interpretation: Clarity is a strong premium factor independent of size.

#### 3. Premium gap widens as carat increases

At low carat (<0.5), price differences across clarity grades are relatively small.

At higher carats (>1.5), price differences between clarity grades become much wider.

Interpretation: Clarity matters more for large diamonds than small ones.

#### 4. IF and VVS grades show the highest price curves

IF (Internally Flawless) and VVS1/VVS2 have the steepest and highest smooth curves.

These diamonds show strong price appreciation with increasing carat.

Interpretation: High-clarity diamonds scale better in value as size increases.

#### 5. SI1 and SI2 lie in the mid-price segment

SI1 and SI2 diamonds form a middle band between premium (VVS/IF) and low clarity (I1).

Their growth trend is positive but less steep.

Interpretation: These categories offer a balance between affordability and quality.

#### 6. I1 clarity diamonds are consistently discounted

I1 diamonds show lowest prices across all carat sizes.

Even at higher carats, I1 does not catch up with higher clarity grades.

Interpretation: Poor clarity strongly limits price potential, regardless of size.

#### 7. Price variability increases with carat

At higher carats, points are more spread out vertically.

This indicates greater price dispersion due to combined effects of clarity, cut, and other attributes.

Interpretation: Large diamonds have more complex pricing dynamics.

### Business Insights
##### 1.Carat size is the main driver of revenue
Prices increase sharply with carat size, especially for larger diamonds.

Decision: Prioritize sourcing and marketing higher-carat diamonds to drive overall revenue growth.

##### 2.Clarity strongly influences pricing at the same carat
Higher clarity diamonds consistently command higher prices than lower clarity diamonds of equal size.

Decision: Use clarity-based pricing tiers and clearly differentiate products by clarity grade.

##### 3.Clarity impact increases for diamonds above one carat
Price gaps between clarity grades widen significantly for larger diamonds.

Decision: For diamonds above one carat, focus on higher clarity grades to maximize returns.

##### 4.High-clarity diamonds deliver superior margins
IF and VVS diamonds show faster price appreciation compared to other grades.

Decision: Position IF and VVS diamonds as premium offerings and allocate more inventory budget to these categories.

##### 5.Mid-clarity diamonds support volume sales
SI1 and SI2 diamonds offer a balance of price and quality and likely generate higher sales volumes.

Decision: Maintain strong inventory levels of SI1 and SI2 diamonds to ensure stable cash flow.

##### 6.Low-clarity diamonds serve price-sensitive customers
I1 clarity diamonds remain consistently low-priced, even at higher carats.

Decision: Limit exposure to large I1 diamonds and use them mainly as entry-level products.

##### 7.Diamond pricing follows a non-linear pattern
Price does not increase evenly with carat; it depends on both carat and clarity together.

Decision: Implement non-linear pricing models and optimize inventory mix based on combined carat-clarity analysis.

### Conclusion
Overall, the analysis shows that carat size drives demand and revenue, while clarity determines price differentiation and profitability. The impact of clarity becomes significantly stronger for larger diamonds, making high-clarity, high-carat stones the key contributors to premium margins. Mid-clarity diamonds support stable sales volumes, while low-clarity diamonds primarily serve price-sensitive customers. Therefore, effective business decisions should focus on non-linear pricing strategies, optimized inventory allocation, and clear market segmentation to balance revenue growth, profitability, and risk.